---
title: "ST558 project 2"
author: "Stephanie Shuai"
format: html
editor: visual
---

===============================================================

# NBA Shot Data Analysis Project

===============================================================

# Loading Packages

We begin by loading the tidyverse package. The tidyverse is a collection of cohesive R packages. It automatically loads essential packages such as dplyr for data manipulation, readr for data import, and ggplot2 for data visualization.

```{r}
library(tidyverse)
library(dplyr)
library(readr)
library(ggplot2)
library(ggstatsplot)
```

------------------------------------------------------------------------

# Data Import and Data Structure Processing

------------------------------------------------------------------------

## Read NBA shooting data from a CSV file

Next, we imported the NBA play-by-play dataset into R for using read_csv() from the readr package.

```{r}
NBA_data <- read_csv("~/ST558_Project2/NBA/NBA_PBP_2019-20.csv")
```

## Previewing Data

To verify that the dataset has been imported correctly and to perform a quick visual inspection of the column alignment, we print the first six rows of the dataset using the head() function.

```{r}
head(NBA_data)
```

## Data Structure

Then, we inspect the structural dimensions and variable data types of the dataset using glimpse(). It transposes the data presentation, displaying each column vertically along with its type and a preview of the initial values. This provides a highly readable summary of the datasets's columns.

```{r}
glimpse(NBA_data)
```

## Checking Unique Values

Before performing data cleaning and factor conversion, we explore the distinct levels within the categorical variables. Using a pipeline, we select key variables of interest, and apply the unique() function to each column via map() form the purrr package.

```{r}
NBA_data %>%
  select("GameType", "WinningTeam", "Quarter", "AwayTeam", "HomeTeam",
  "ShotType", "ShotOutcome", "FoulType", "ReboundType",
  "ViolationType", "FreeThrowOutcome", "FreeThrowNum",
  "TurnoverType", "TurnoverCause"
  ) %>%
  map(unique)
```

## Data Cleaning

In this step, we clean and preprocess the raw dataset to ensure all variables are in the correct format for statistical modeling.

First, we use 'na.if()' across all character columns to replace the empty string entries with formal 'NA' missing values.

Next, we convert categorical variables into R 'factor' objects and explicitly specify their levels.

Finally, we convert game score, time remaining, and shot distance variables.

```{r}
NBA_data <- NBA_data %>%
  mutate(
    # Convert empty strings to NA
    across(where(is.character), ~ na_if(., "")),
    
    # Factor levels
    GameType = factor(GameType, levels = c("playoff", "regular")),
    WinningTeam = factor(WinningTeam, levels = c(
      "TOR", "LAC", "CHO", "DET", "ORL", "MIN", "MIA", "PHI", "DAL", "SAS",
      "UTA", "PHO", "DEN", "ATL","MIL", "BOS", "BRK", "CHI", "WAS", "POR",
      "LAL", "CLE", "HOU", "OKC", "MEM", "NYK", "GSW", "IND", "NOP","SAC"
      )),
   Quarter = factor(Quarter, levels = c(1, 2, 3, 4, 5, 6)),
   AwayTeam = factor(AwayTeam, levels = c(
     "NOP", "LAL", "CHI", "DET", "CLE", "MIN","MEM", "BOS", "WAS", "NYK", "OKC", "SAC",
     "DEN", "ATL", "MIL", "LAC", "TOR", "DAL","PHO", "POR", "UTA", "MIA", "PHI", "ORL",
     "IND", "GSW", "BRK", "CHO", "HOU", "SAS"
     )),
   HomeTeam = factor(HomeTeam, levels = c(
     "TOR", "LAC", "CHO", "IND", "ORL", "BRK", "MIA", "PHI", "DAL", "SAS", "UTA",
     "PHO", "POR", "DET", "HOU", "GSW", "BOS", "MEM", "NOP", "OKC", "DEN", "SAC",
     "LAL", "MIL", "ATL", "NYK", "CHI", "CLE", "MIN", "WAS" 
     )),
   ShotType = factor(ShotType, 
                          levels = c( "2-pt jump shot",
                                      "2-pt layup",
                                      "3-pt jump shot",
                                      "2-pt hook shot",
                                      "2-pt dunk")),
   ShotOutcome = factor(ShotOutcome, levels = c("miss","make")),
   FoulType = factor(FoulType,levels = c(
     "shooting", "offensive", "personal", "loose ball","technical", 
     "away from play","personal take", "flagrant", "clear path"
     )),
   FreeThrowOutcome = factor(FreeThrowOutcome, levels = c("make", "miss")),
   FreeThrowNum = factor(FreeThrowNum, levels = c(
    "1 of 2", "2 of 2", "1 of 1", "1 of 3", "2 of 3", "3 of 3", "technical"
     )),
   ViolationType = factor(ViolationType, levels = c(
     "kicked ball", "def goaltending", "delay of game", "jump ball", "double lane",
      "violation"
     )),
  TimeoutTeam = factor(TimeoutTeam, levels = c(
    "TOR", "NOP", "LAC", "LAL", "CHO", "CHI", "IND", "DET", "ORL", "CLE",
    "BRK", "MIN", "MEM", "MIA", "PHI", "BOS", "DAL", "WAS", "NYK", "SAS", "UTA",
    "OKC", "PHO", "SAC", "POR", "DEN", "ATL", "MIL", "HOU", "GSW"
    )),
  TurnoverType = factor(TurnoverType, levels = c(
      "offensive foul", "bad pass", "traveling", "out of bounds lost ball", "8 sec",
      "lost ball", "step out of bounds", "back court", "3 sec", "shot clock", "palming",
      "off goaltending", "dbl dribble", "illegal assist", "turnover", "lane violation",
      "discontinued dribble", "inbound", "5 sec"
      )),
  TurnoverCause = factor(TurnoverCause, levels = c("steal")),
  ReboundType = factor( ReboundType, levels = c("offensive", "defensive")),
  
  # Convert numerical columns
  SecLeft = as.numeric(SecLeft),
  AwayScore = as.numeric(AwayScore),
  HomeScore = as.numeric(HomeScore),
  ShotDist = as.numeric(ShotDist)
  ) 

```

------------------------------------------------------------------------

# Descriptive Statistical Analysis

------------------------------------------------------------------------

## One-way Contingency tables

To explore the baseline frequency distribution of individual categorical variables in the dataset, we construct one-way contingency tables using the table function. These tables summarize the counts of observations within each category, helping us understand the distribution of game types, shot types, shot outcomes, fouls, quarters, turnovers, and violations.

```{r}
# Frequency table for game type
table(NBA_data$GameType)

# Frequency table for Shot type
table(NBA_data$ShotType)

# Frequency table for Shot Outcome
table(NBA_data$ShotOutcome)

# Frequency table for Foul type
table(NBA_data$FoulType)

# Frequency table for Quarter
table(NBA_data$Quarter)

# Frequency table for Turnover Type
table(NBA_data$TurnoverType)

# Frequency table for Violation Type
table(NBA_data$ViolationType)
```

## Two-way Contingency tables

To analyze the relationships and potential associations between pairs of categorical variables, we construct two-way contingency tables. Cross-tabulating these variables allows us to observe patterns and trends, such as how shot tyeps relate to shout outcomes, or how foul types vary by game type.

```{r}
# Game type vs. Shot outcome
table(NBA_data$GameType, NBA_data$ShotOutcome)

# Shot type vs. Shot outcome
table(NBA_data$ShotType, NBA_data$ShotOutcome)

# Quarter vs. Shot outcome
table(NBA_data$Quarter,NBA_data$ShotOutcome)

# Game type vs. Foul type
table(NBA_data$GameType, NBA_data$FoulType)

# Quarter vs. Turnover type
table(NBA_data$Quarter, NBA_data$TurnoverType)

# Game type vs. Quarter
table(NBA_data$GameType, NBA_data$Quarter)

# Quarter vs. Foul type
table(NBA_data$Quarter, NBA_data$FoulType)

```

# Numerical summaries

## Filter Missing Values and Summarize Shot Distance by Outcome

Before computing summary statistics, we filter the dataset to create 'NBA_data_clean', removing any rows with missing values ('NA') in our primary analysis variables: 'ShotOutcome', 'ShotDist', and 'GameType'.

We then calculate numerical summaries for the 'ShotDist'(Shot Distance) variable, grouped by the 'ShotOutcome'(make vs. miss). The summary metrics include the sample size, mean, median, standard deviation, minimum, and maximum. This helps us examine whether the average distance of a shot differs between successful and unsuccessful attempts.

```{r}
# Create clean data set
NBA_data_clean <- NBA_data  %>%
  filter(!is.na(ShotOutcome),
         !is.na(ShotDist),
         !is.na(GameType)) 

# Numerical Summary
NBA_data_clean %>%
  group_by(ShotOutcome)  %>%
  summarise(
    n = n(),
    mean = mean(ShotDist, na.rm = TRUE),
    median = median(ShotDist, na.rm = TRUE),
    sd = sd(ShotDist, na.rm = TRUE),
    min = min(ShotDist, na.rm = TRUE),
    max = max(ShotDist, na.rm = TRUE)
  )
```

## Summarize Away Team Score by Game Type

To explore if the offensive output of visiting teams varies between different stages of the season, we group the cleaned data by GameType(playoff vs. regular). We then compute the descriptive statistics for the AwayScore variable.

```{r}
NBA_data_clean %>%
  group_by(GameType) %>%
  summarise(
    n = n(),
    mean = mean(AwayScore, na.rm = TRUE),
    median = median(AwayScore, na.rm = TRUE),
    sd = sd(AwayScore, na.rm = TRUE),
    min = min(AwayScore, na.rm = TRUE),
    max = max(AwayScore, na.rm = TRUE)
  )
```

```{r}
NBA_data_clean %>%
  group_by(Quarter) %>%
  summarise(
    n = n(),
    mean = mean(SecLeft, na.rm = TRUE),
    median = median(SecLeft, na.rm = TRUE),
    sd = sd(SecLeft, na.rm = TRUE),
    min = min(SecLeft, na.rm = TRUE),
    max = max(SecLeft, na.rm = TRUE)
  )
```

------------------------------------------------------------------------

# Data Visualization

------------------------------------------------------------------------

we present six distinct visualizations.

## Plot 1: Bar chart(Shot Types)

To examine the frequency and popularity of different shot selections, we construct a bar chart of Shot Types. The height of each bar represents the absolute count of shots attempted for each category.

```{r}
ggplot(NBA_data_clean, aes(x = ShotType)) + 
  geom_bar() +
  labs(
    title = "Counts of Shot Types",
    subtitle = "2019-20 NBA Play-by-Play Data",
    x = "Shot Types",
    y = "Count"
  ) + 
  theme_minimal() +
  theme(legend.position = "none")
```

## Plot 2: Density Plot (Shot Distance)

To visualize the continuous probability distribution of shot distances, we construct a kernel density plot using geom_density(). We fill the area under the curve with a custom color and apply a semitransparent opacity to achieve a modern look. The plot shows where players are most active on the court.

```{r}
ggplot(NBA_data_clean, aes(x = ShotDist)) +
  geom_density(fill = "#1e3d59", alpha = 0.8 ) +
  labs(
    title = "Density of Shot Distance",
    x = "Shot Distance(feet)",
    y = "Density"
  ) + theme_minimal()
```

## Plot 3: Side-by-side Bar Chart (Multivariate - Shot Outcomes by Game Type)

To investigate whether players' shot outcomes differ between high-pressure playoff games compared to regular season games, we create a side-by-side bar chart. We use fill = GameType to color the bars by game type and set position = "dodge" to place the playoff and regular season bars next to each other for easy comparison.

```{r}
ggplot(NBA_data_clean, aes(x = ShotOutcome, fill = GameType)) +
  geom_bar(position = "dodge", alpha = 0.9) +
  scale_fill_manual(values = c("playoff" = "#ff6e40", "regular" = "#1e3d59")) +
  labs(
    title = "Counts of Shot Outcomes by Game Type",
    x = "Shot Outcome",
    y = "Count",
    fill = "Game Type"
  ) 
```

## Plot 4: Box Plot (Multivariate - Shot Distance by Shot Outcomes)

To analyze the spread and key statistics of shot distances grouped by shot success, we construct side-by-side box plots. This visualization helps us assess the relationship between how far a shot is taken and whether it is successfully made.

```{r}
ggplot(NBA_data_clean , aes(x = ShotOutcome, y = ShotDist, fill = ShotOutcome)) +
  geom_boxplot()+
  scale_fill_manual(values = c("miss" = "#e85a4f", "make" = "#45b6fe")) +
  labs(
    title = "Shot Distance by Shot Outcome",
    x = "Shot Outcome",
    y = "Shot Distance(feet)"
  )
```

## Plot 5: Faceted Histogram (Multivariate & Faceted - Shot Distance by Quarter)

To analyze how shot distance and success rates evolve as the game progresses, we construct a faceted histogram.

```{r}
ggplot(NBA_data_clean, aes(x = ShotDist, fill = ShotOutcome)) +
  geom_histogram(bins = 20, position = "identity", alpha = 0.6) +
  facet_wrap(~ Quarter, labeller = label_both) +
  scale_fill_manual(values = c("miss" = "#e85a4f", "make" = "#45b6fe")) +
  labs(
    title = "Density of Shot Distance by Quarter and Outcome",
    x = "Shot Distance(feet)",
    y = "Density",
    fill = "Shot Outcome"
  )
```

## Plot 6: Scatter Plot (Multivariate - Shot Distance vs. Seconds Left)

To investigate the relationship between the time left on the clock in a quarter and the shot distance, we construct a scatter plot.

```{r}
ggplot(NBA_data_clean, aes(x = SecLeft, y = ShotDist, color = ShotOutcome)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("miss" = "#e85a4f", "make" = "#45b6fe")) +
  labs(
    title = "Shot Distance vs. Seconds Left in Quarter",
    x = "Seconds Remaining in Quarter",
    y = "Shot Distance(feet)",
    color = "Shot Outcome"
  ) 
```

## Plot 7: Bivariate Highest Density Region (HDR) Plot

To investigate whether shot distances differ significantly among the top three most frequent shot types, we construct a categorical statistical comparison plot using ggbetweenstats().

```{r}
top3_shot_types <- NBA_data_clean %>% 
  dplyr::count(ShotType, sort = TRUE) %>%
  slice_head(n = 3) %>%
  pull(ShotType)

NBA_top3 <- NBA_data_clean %>% 
  filter(ShotType %in% top3_shot_types)

if (nrow(NBA_top3) > 300){
  set.seed(123)
  NBA_top3 <- NBA_top3[sample(1:nrow(NBA_top3), 300), ]
}

ggbetweenstats(
  data  = NBA_top3,
  x     = ShotType, 
  y     = ShotDist,
  title = "Shot Distance by Shot Type(Top3)",
  xlab  = "Shot Type",
  ylab  = "Shot Distance(feet)",
  type  = "nonparametric",
  pairwise.comparisons = TRUE, 
  p.adjust.method = "BH"
)
```
